---
title: "Tidy Data"
author: "RNepal"
date: "9/16/2019"
output: html_document
---


<!-- This file is provided in Data Analysis with R bootcamp by R Nepal (www.rugnepal.org) -->

# Tidy Data Principle


1. Each variable is in its own column 
2. Each observation is in its own row
3. Each value is in its own cell

# Creating a Data Frame

```{r}
# data frame with 4 cols, 3 rows
data <- data.frame(id = c(1, 2), 
                   x = c("a", "b"),
                   y = c("c", "d"),
                   z =  c("e", "f"))

data
```

# Pivoting - Wide to Long

```{r}
library(tidyr)

long <- pivot_longer(data,
  cols =2:4 ,
  names_to = "key",
  values_to = "val"
)

long

```


# Pivoting - Long to Wide

```{r}
wide <- pivot_wider(long,
  names_from =key ,
  values_from =val 
)

wide
```

# table4a - Reshaping Data

```{r}

table_a <- table4a %>% 
  pivot_longer(cols =2:3, 
              names_to = "year ", 
              values_to = "cases")

table_a
```

# table4b-Reshaping data

table4b - Reshape data into tidy format


```{r}
table_b<-table4b %>%
  pivot_longer(cols=2:3,
               names_to ="year",
               values_to="population")

table_b



table_b
```

# Merging Data

```{r}
library(tidyr)
library(dplyr)
df <- merge(table_a,table_b)



df
```

```{r}
library(tidyr)
library(dplyr)
df <- merge(table_a, table_b) 
df %>% 
  filter(year==2000) %>% 
  select(-year) %>% 
  mutate(rate=cases/population) %>% 
  summarise(avg=mean(rate))



```


***

# Take Aways

Data comes in many formats but R prefers just one: _tidy data_.

A data set is tidy if and only if:

1. Every variable is in its own column
2. Every observation is in its own row
3. Every value is in its own cell (which follows from the above)

What is a variable and an observation may depend on your immediate goal.
